---
title: "Exploratory Data Analysis (EDA)"
output: pdf_document
date: "2026-03-11"
---

<!--
================================================================================
01_data_exploration.Rmd

PURPOSE:
  This notebook performs exploratory data analysis (EDA) on FDIC Call Report
  data (balance sheets and income statements). It serves as the first step in
  the bank financial-deterioration prediction pipeline.

WORKFLOW POSITION:
  1. **01_data_exploration.Rmd** ← (this file)
  2. 02_data_cleaning.Rmd
  3. 03_data_preparation.Rmd
  4. Model training scripts (XGBoost, LightGBM, Random Forest, etc.)

INPUTS:
  - ../data/raw_data/call-reports-balance-sheets-Jan2026.dta
  - ../data/raw_data/call-reports-income-statements-Jan2026.dta

OUTPUTS:
  - ../data/processed_data/df.rds          — merged & lightly trimmed panel
  - ../data/processed_data/bank_panel.rds  — per-bank quarter counts

KEY ANALYSES:
  - Data loading, column pruning, and merging of balance-sheet / income-
    statement datasets
  - Panel dimensions: number of unique banks and reporting periods
  - Temporal coverage and bank entry/exit patterns
  - Bank lifespan distribution
  - Missingness profile across all variables
  - Quarterly seasonality in key balance-sheet items (assets, loans, deposits,
    equity)
================================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.path = "../results/figures/01_exploration-"
)
```

## 1. Load Required Libraries

```{r load-libraries}
# tidyverse  — data wrangling (dplyr, tidyr, ggplot2, etc.)
# lubridate  — date manipulation helpers
# skimr      — quick summary statistics
# janitor    — data-cleaning utilities (clean_names, tabyl, etc.)
# haven      — read Stata (.dta) files
# readxl     — read Excel (.xlsx) files
library(tidyverse)
library(lubridate)
library(skimr)
library(janitor)
library(haven)
library(readxl)
```

---

## 2. Load and Inspect the Dataset

Load the two core FDIC Call Report files — **balance sheets** and **income
statements** — then drop granular or redundant columns that are not needed for
this analysis.

```{r load-data}
# Read Stata (.dta) files produced by the FDIC RSSD data pipeline
df_bs <- read_dta("../data/raw_data/call-reports-balance-sheets-Jan2026.dta")
df_is <- read_dta("../data/raw_data/call-reports-income-statements-Jan2026.dta")
```

### 2.1 Column Selection — Drop Redundant / Granular Variables

Many of the original variables are highly granular breakdowns (e.g., time
deposits by size bucket, maturity band, or average balances) that would add
noise without clear value at this stage. We retain the broad aggregates and
drop the detailed sub-components.

```{r drop-columns}
# ---- Balance-sheet columns to drop ----
# Categories removed:
#   • Detailed deposit breakdowns (by size, maturity, domestic/foreign)
#   • Quarterly average balances (qtr_avg_*)
#   • Maturity-band breakdowns for securities, loans & leases
#   • Brokered-deposit detail
#   • Interest-rate derivative notional amounts
#   • Administrative / geographic identifiers
df_bs <- df_bs %>%
  select(-c(
    transaction_dep,
    time_dep_ge100k,
    time_cd_ge100k,
    time_dep_lt100k,
    time_dep_gt250k,
    time_ge100k_le250k,
    time_dep_le250k,
    securities_ac,
    htmsec_ac,
    afssec_fv,
    domestic_dep,
    foreign_dep,
    dom_deposit_ib,
    dom_deposit_nib,
    for_deposit_ib,
    for_deposit_nib,
    time_sav_dep,
    nontransaction_dep,
    uninsured_time_dep,
    nontransaction_sav_dep,
    tot_sav_dep,
    qtr_avg_ib_bal_due,
    qtr_avg_securities,
    qtr_avg_ust_sec,
    qtr_avg_mbs,
    qtr_avg_oth_sec,
    qtr_avg_trad_ass,
    qtr_avg_ffrepo_ass,
    qtr_avg_ln_tot,
    qtr_avg_ln_re,
    qtr_avg_ln_rre,
    qtr_avg_ln_othre,
    qtr_avg_ln_agr,
    qtr_avg_ln_ci,
    qtr_avg_ln_cons,
    qtr_avg_ln_cc,
    qtr_avg_ln_othcons,
    qtr_avg_ln_fgn,
    qtr_avg_lease,
    qtr_avg_assets,
    qtr_avg_trans_dep_dom,
    qtr_avg_sav_dep_dom,
    qtr_avg_time_dep_ge100k,
    qtr_avg_time_dep_gt250k,
    qtr_avg_time_dep_lt100k,
    qtr_avg_time_dep_le250k,
    qtr_avg_time_cd_ge100k,
    qtr_avg_fgn_dep,
    qtr_avg_ffrepo_liab,
    qtr_avg_othbor_liab,
    ust_sec_1y_3y,
    ust_sec_3y_5y,
    ust_sec_5y_15y,
    ust_sec_over_15y,
    res_mbs_1y_3y,
    res_mbs_3y_5y,
    res_mbs_5y_15y,
    res_mbs_over_15y,
    other_mbs_3y_less,
    other_mbs_3y_more,
    securities_1y_3y,
    securities_3y_5y,
    securities_5y_15y,
    securities_over_15y,
    res_loans_1y_3y,
    res_loans_3y_5y,
    res_loans_5y_15y,
    res_loans_over_15y,
    ln_lease_1y_3y,
    ln_lease_3y_5y,
    ln_lease_5y_15y,
    ln_lease_over_15y,
    brokered_dep_lt100k,
    brokered_dep_ge100k,
    time_dep_lt100k_1y_3y,
    time_dep_lt100k_over_3yr,
    time_dep_ge100k_1y_3y,
    time_dep_ge100k_1y_3y,
    time_dep_ge100k_over_3yr,
    time_dep_le250k_1y_3y,
    time_dep_le250k_over_3yr,
    time_dep_gt250k_1y_3y,
    time_dep_gt250k_over_3yr,
    ust_sec_3mo_less,
    ust_sec_3mo_12mo,
    res_mbs_3mo_less,
    res_mbs_3mo_12mo,
    securities_3mo_less,
    securities_3mo_12mo,
    securities_mat_1y_less,
    res_loans_3mo_less,
    res_loans_3mo_12mo,
    ln_lease_3mo_less,
    ln_lease_3mo_12mo,
    ln_lease_mat_1y_less,
    time_dep_lt100k_3mo_less,
    time_dep_lt100k_3mo_12mo,
    time_dep_lt100k_mat_1y_less,
    time_dep_ge100k_3mo_less,
    time_dep_ge100k_3mo_12mo,
    time_dep_le250k_3mo_less,
    time_dep_le250k_3mo_12mo,
    time_dep_le250k_mat_1y_less,
    time_dep_gt250k_3mo_less,
    time_dep_gt250k_3mo_12mo,
    tot_gna_deriv_ir,
    tot_gna_deriv_ir_par,
    tot_gna_ir_fixed_rate_swap,
    tot_swaps_int_rate,
    gross_hedging,
    tot_gna_ir_float_rate_swap,
    net_hedging,    
    entity_type,
    fgn_call_cntry_cd,
    fgn_call_fam_id,
    chtr_type_cd,
    nm_short,
    act_prim_cd,
    zip_cd,
    city,
    county_cd,
    dist_frs,
    cntry_cd))

# ---- Income-statement columns to drop ----
# Categories removed:
#   • Detailed interest income by loan type (real estate, agriculture, etc.)
#   • Detailed interest expense by deposit type and maturity
#   • Trading revenue and dividend sub-items
df_is <- df_is %>%
  select(-c(
    ytdint_exp_time_ge100k_dom,
    ytdint_exp_time_lt100k_dom,
    ytdint_exp_time_le250k_dom,
    ytdint_exp_time_gt250k_dom,
    ytdint_exp_time_dep,
    ytdtradrev_inc,
    ytdcommdividend,
    ytdint_inc_sec_ust,
    ytdint_inc_sec_mbs,
    ytdint_inc_sec_oth,
    ytdint_inc_ln_re,
    ytdint_inc_ln_rre,
    ytdint_inc_ln_othre,
    ytdint_inc_ln_agr,
    ytdint_inc_ln_ci,
    ytdint_inc_ln_cc,
    ytdint_inc_ln_othcons,
    ytdint_inc_ln_fgn,
    ytdint_inc_ln_indiv,
    ytdint_exp_trans_dep_dom,
    ytdint_exp_savings_dep_dom,
    ytdint_exp_fgn))
```

### 2.2 Merge Balance-Sheet and Income-Statement Data

Join the two datasets on `id_rssd` (unique bank identifier) and `date`
(quarter index) to produce a single unified panel.

```{r merge-bs-is}
df <- df_bs %>%
  left_join(
    df_is,
    by = c("id_rssd", "date")
  )
```

---

## 3. Panel Dimensions

```{r unique-counts}
# Count the number of distinct banks and reporting periods in the panel
n_banks   <- n_distinct(df$id_rssd)
n_periods <- n_distinct(df$date)

cat("Unique banks:  ", n_banks, "\n")
cat("Unique periods:", n_periods, "\n")
```

---

## 4. Temporal Coverage

### 4.1 Observations per Quarter

Plot the number of bank-quarter observations over time to identify any
reporting gaps or structural breaks in the data.

```{r year-coverage}
df %>%
  count(date) %>%
  ggplot(aes(date, n)) +
  geom_line() +
  labs(title="Observations per Year",
       x="Year",
       y="Number of Bank-Quarter Observations")
ggsave("../results/figures/01_obs_per_year.png", width=8, height=5)
```

### 4.2 Create Readable Date Variables

The raw `date` column is an integer encoding where the origin is 1960-Q1.
Convert it to explicit `year`, `qtr`, and a proper calendar date (`qtr_date`)
for cleaner plotting and aggregation.

```{r create-year}
df <- df %>%
  mutate(
    year = 1960 + floor(date / 4),              # Extract year from encoded date
    qtr  = (date %% 4) + 1,                     # Extract quarter (1–4)
    year_quarter = paste0(year, " Q", qtr),      # Human-readable label
    qtr_date = make_date(year, qtr * 3, 1) +     # Calendar end-of-quarter date
               months(1) - days(1)
  )
```

### 4.3 Active Banks per Year

Visualise how many distinct banks file reports each year. A declining trend
may reflect industry consolidation.

```{r banks-per-year}
df %>%
  group_by(year) %>%
  summarise(banks = n_distinct(id_rssd)) %>%
  ggplot(aes(year, banks)) +
  geom_line() +
  labs(title="Number of Active Banks per Year",
       x="Year",
       y="Banks")
ggsave("../results/figures/01_active_banks_per_year.png", width=8, height=5)
```

---

## 5. Bank Lifespan Analysis

### 5.1 Compute Lifespan per Bank

For each bank, compute its first and last observed year, total years active,
and the number of quarterly observations on file.

```{r bank-lifespan}
bank_lifespan <- df %>%
  group_by(id_rssd) %>%
  summarise(
    first_year   = min(year),
    last_year    = max(year),
    years_active = last_year - first_year + 1,
    quarters_obs = n()
  ) %>%
  arrange(desc(years_active))

head(bank_lifespan)
```

### 5.2 Lifespan Distribution

```{r lifespan-distribution}
bank_lifespan %>%
  ggplot(aes(years_active)) +
  geom_histogram(bins = 40) +
  labs(title="Distribution of Bank Lifespans",
       x="Years Active",
       y="Number of Banks")
```

### 5.3 Lifespan Frequency Table

```{r lifespan-count}
bank_lifespan %>%
  count(years_active) %>%
  arrange(years_active)
```

### 5.4 Panel Balance Check

Summarise the distribution of the number of quarters observed per bank.
A perfectly balanced panel would show a single value.

```{r panel-balance}
bank_panel <- df %>%
  group_by(id_rssd) %>%
  summarise(quarters = n())

summary(bank_panel$quarters)
```

---

## 6. Bank Entry and Exit Patterns

### 6.1 Entry (First Observation Year)

```{r entry-exit}
bank_lifespan %>%
  ggplot(aes(first_year)) +
  geom_histogram(bins = 50) +
  labs(title="Bank Entry Over Time")
```

### 6.2 Exit (Last Observation Year)

```{r exits}
bank_lifespan %>%
  ggplot(aes(last_year)) +
  geom_histogram(bins = 50) +
  labs(title="Bank Exit Over Time")
```

---

## 7. Missingness Profile

Compute the fraction of missing values for every column and display the 50
worst offenders. High missingness may indicate variables that are only
collected in certain periods or for certain charter types.

```{r missing}
df %>%
  summarise(across(everything(), ~mean(is.na(.)))) %>%
  pivot_longer(everything()) %>%
  arrange(desc(value)) %>%
  head(50)
```

---

## 8. Seasonality Analysis — Assets, Loans, Deposits, Equity

Investigate whether key balance-sheet items show systematic quarterly
patterns (e.g., window-dressing effects at year-end).

### 8.1 Subset Key Variables

```{r df-key-vars}
# Create a focused dataframe with only the key balance-sheet items
df_key <- df %>%
  select(id_rssd, year, qtr, assets, ln_tot, deposits, equity) %>%
  rename(loans = ln_tot) %>%
  mutate(qtr_label = factor(paste0("Q", qtr), levels = paste0("Q", 1:4)))
```

### 8.2 Aggregate Medians by Year-Quarter

```{r aggregate-by-quarter}
# Aggregate: median across all banks for each year-quarter
# Median is robust to outliers created by very large / very small banks
df_key_agg <- df_key %>%
  group_by(year, qtr, qtr_label) %>%
  summarise(
    assets   = median(assets,   na.rm = TRUE),
    loans    = median(loans,    na.rm = TRUE),
    deposits = median(deposits, na.rm = TRUE),
    equity   = median(equity,   na.rm = TRUE),
    .groups  = "drop"
  )
```

### 8.3 Quarterly Seasonality Box-Plots

Box-plots of median values grouped by quarter. Differences in median or
spread across quarters would indicate seasonality.

```{r seasonality-boxplots, fig.width=10, fig.height=8}
# Pivot to long form for faceted plotting
df_key_long <- df_key_agg %>%
  pivot_longer(cols = c(assets, loans, deposits, equity),
               names_to  = "variable",
               values_to = "value") %>%
  mutate(variable = factor(variable,
                           levels = c("assets", "loans", "deposits", "equity"),
                           labels = c("Assets", "Loans", "Deposits", "Equity")))

# Box-plots: distribution of median values by quarter
ggplot(df_key_long, aes(x = qtr_label, y = value / 1e3, fill = qtr_label)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(title    = "Quarterly Seasonality — Key Balance-Sheet Variables",
       subtitle = "Distribution of yearly median values by quarter (across all years)",
       x = "Quarter", y = "Value (thousands $)",
       fill = "Quarter") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

### 8.4 Median Time-Series Coloured by Quarter

Overlay quarterly time-series to visually detect persistent gaps between
quarters (a hallmark of seasonality).

```{r seasonality-lines, fig.width=10, fig.height=8}
# Line chart: median over time coloured by quarter
ggplot(df_key_long, aes(x = year, y = value / 1e3, colour = qtr_label, group = qtr_label)) +
  geom_line(alpha = 0.8, linewidth = 0.6) +
  geom_point(size = 0.8, alpha = 0.6) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_colour_brewer(palette = "Set1") +
  labs(title    = "Median Values Over Time by Quarter",
       subtitle = "Each line represents one quarter — gaps between lines suggest seasonality",
       x = "Year", y = "Value (thousands $)",
       colour = "Quarter") +
  theme_minimal(base_size = 12)
```

---

## 9. Save Intermediate Outputs

Persist the merged panel (`df`) and the bank-level quarter counts
(`bank_panel`) for use in the next notebook (`02_data_cleaning.Rmd`).

```{r save-outputs}
saveRDS(df, "../data/processed_data/df.rds")
saveRDS(bank_panel, "../data/processed_data/bank_panel.rds")
```


