---
title: "Data Cleaning"
output:
  pdf_document: default
  html_document: default
date: "2026-03-29"
---

<!--
================================================================================
02_data_cleaning.Rmd

PURPOSE:
  Clean the merged Call Report panel from 01_data_exploration.Rmd. Remove
  non-U.S. banks, young/short-lived banks, and tiny banks. Engineer intra-year
  volatility features, then collapse to annual (Q4) snapshots.

INPUTS:
  - ../data/processed_data/df.rds, bank_panel.rds

OUTPUTS:
  - ../data/processed_data/df_clean.rds — cleaned annual panel (Q4 only)

CLEANING STEPS:
  1. Keep only U.S.-domiciled banks
  2. Drop each bank's first 3 years (start-up phase)
  3. Remove banks with < 17 quarters of data
  4. Engineer intra-year features (QoQ changes, within-year SD, max decline)
  5. Collapse to Q4 (year-end), drop partial 2025
  6. Winsorise bottom 1% of assets per year
  7. Deduplicate bank-year ties
================================================================================
-->

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.path = "../results/figures/02_cleaning-"
)

# Initialize attrition tracking
attrition <- tibble(Step = "Raw Data", Banks = 0, Observations = 0)
```

## 1. Load Libraries

```{r load-libraries}
library(lubridate)
library(skimr)
library(janitor)
library(haven)
library(readxl)
```

## 2. Load Input Data

```{r load-data}
# Merged panel and bank-level quarter counts from 01_data_exploration.Rmd
df         <- readRDS("../data/processed_data/df.rds")
bank_panel <- readRDS("../data/processed_data/bank_panel.rds")

attrition <- bind_rows(attrition, 
  tibble(Step = "Initial Merge", 
         Banks = n_distinct(df$id_rssd), 
         Observations = nrow(df)))
```

---

## 3. Geographic Filter — U.S. Banks Only

Foreign-chartered banks follow different regulatory standards; exclude them.

```{r filter-us}
df <- df %>%
  filter(cntry_nm == "UNITED STATES")
```

## 4. Drop Start-Up Phase Observations

Exclude each bank's first 3 years to avoid start-up growth artefacts.

```{r drop-startup}
df <- df %>%
  mutate(
    open_year = year(ymd(dt_open))      # Parse bank opening date
  ) %>%
  filter(
    year >= open_year + 3               # Keep data >= 3 yrs post-opening
  )
```

## 5. Remove Short-Lived Banks

Banks with < 17 quarters (~4 years) lack sufficient history for rolling
features and are likely data artefacts.

```{r drop-short-lived}
short_lived_banks <- bank_panel %>%
  filter(quarters < 17) %>%
  pull(id_rssd)

# Drop them
df <- df %>%
  filter(!id_rssd %in% short_lived_banks)
```

---

## 6. Engineer Intra-Year (Quarterly) Features

Compute QoQ changes within each bank-year, then summarise quarterly data
into bank-year-level volatility and stress measures:
- `*_sd_intra`    — within-year std. dev. of each metric
- `*_avg_yr`      — within-year mean level
- `*_max_decline` — worst single-quarter drop

Only bank-years with all 4 quarters are retained.

```{r intra-year-features}
library(data.table)

dt <- as.data.table(df)
setorder(dt, id_rssd, year, qtr)

# Step 1: QoQ changes via shift() within bank-year
dt[, `:=`(
  assets_qoq   = assets   - shift(assets),
  loans_qoq    = ln_tot   - shift(ln_tot),
  deposits_qoq = deposits - shift(deposits),
  equity_qoq   = equity   - shift(equity)
), by = .(id_rssd, year)]

# Step 2: aggregate per bank-year in one pass
intra_year <- dt[, .(
  n_qtrs = .N,
  assets_sd_intra   = sd(assets,   na.rm = TRUE),
  loans_sd_intra    = sd(ln_tot,   na.rm = TRUE),
  deposits_sd_intra = sd(deposits, na.rm = TRUE),
  equity_sd_intra   = sd(equity,   na.rm = TRUE),
  assets_avg_yr     = mean(assets,   na.rm = TRUE),
  loans_avg_yr      = mean(ln_tot,   na.rm = TRUE),
  deposits_avg_yr   = mean(deposits, na.rm = TRUE),
  equity_avg_yr     = mean(equity,   na.rm = TRUE),
  assets_max_decline   = min(assets_qoq,   na.rm = TRUE),
  loans_max_decline    = min(loans_qoq,    na.rm = TRUE),
  deposits_max_decline = min(deposits_qoq, na.rm = TRUE),
  equity_max_decline   = min(equity_qoq,   na.rm = TRUE)
), by = .(id_rssd, year)][n_qtrs == 4][, n_qtrs := NULL]
```

### 6.1 Join Intra-Year Features to Main Panel

```{r join-intra-year}
df <- df %>%
  left_join(intra_year, by = c("id_rssd", "year"))
```

---

## 7. Collapse to Annual (Q4) Snapshots

Retain only Q4 (year-end) rows — the regulatory reporting standard.
Exclude 2025 as Q4 data may be incomplete.

```{r q4-filter}
# Keep only Q4 (year-end) rows
df <- df %>%
  filter(qtr == 4, year < 2025)
```

## 8. Winsorise Extremely Small Banks

Remove bottom 1% of assets within each year — micro-banks behave
very differently and can distort model estimates.

```{r winsorise-small}
df_clean <- df %>%
  group_by(year) %>%
  mutate(
    assets_p01 = quantile(assets, 0.01, na.rm = TRUE),  # 1st-percentile cutoff
  ) %>%
  ungroup() %>%
  filter(assets >= assets_p01) 
```

## 9. Deduplicate Bank-Year Records

Keep the record with the largest total assets if duplicates exist.

```{r deduplicate}
df_clean <- df_clean %>%
  arrange(id_rssd, year, desc(assets)) %>%
  distinct(id_rssd, year, .keep_all = TRUE)
```

---

## 10. Save Cleaned Dataset

```{r save-clean}
saveRDS(df_clean, "../data/processed_data/df_clean.rds")
```

---

## 11. Cleaning Summary (Attrition Plot)

```{r attrition-plot}
# Final counts
attrition <- bind_rows(attrition, 
  tibble(Step = "Final Cleaned", 
         Banks = n_distinct(df_clean$id_rssd), 
         Observations = nrow(df_clean)))

attrition_plot <- attrition %>%
  mutate(Step = fct_inorder(Step)) %>%
  pivot_longer(cols = c(Banks, Observations), names_to = "Metric", values_to = "Count") %>%
  ggplot(aes(x = Step, y = Count, fill = Metric)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = scales::comma(Count)), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, size = 3) +
  facet_wrap(~ Metric, scales = "free_y") +
  theme_minimal() +
  labs(title = "Data Cleaning Attrition",
       subtitle = "Number of banks and observations remaining after cleaning stages",
       x = NULL, y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(attrition_plot)
ggsave("../results/figures/02_cleaning_attrition.png", plot = attrition_plot, width = 10, height = 6)
```