
title: "Data Preparation"
output:
  pdf_document:
    latex_engine: xelatex
date: "2026-03-17"


<!--
================================================================================
03_data_preparation.Rmd

PURPOSE:
  Build the final modelling dataset from the cleaned annual panel. This
  notebook constructs target variables (crisis indicators), engineers
  financial ratios, rolling statistics, loan/deposit composition features,
  growth rates, and merges macroeconomic control variables.

WORKFLOW POSITION:
  1. 01_data_exploration.Rmd
  2. 02_data_cleaning.Rmd
  3. **03_data_preparation.Rmd** ← (this file)
  4. Model training scripts (XGBoost, LightGBM, Random Forest, etc.)

INPUTS:
  - ../data/processed_data/df_clean.rds     — cleaned annual panel
  - ../data/raw_data/EDOOrders.csv           — FDIC enforcement orders
  - ../data/raw_data/occ_old.xlsx            — OCC old enforcement actions
  - ../data/raw_data/occ_new.csv             — OCC new enforcement actions
  - ../data/raw_data/failures.csv            — FDIC bank failure list
  - ../data/raw_data/macrodata.csv           — World Bank macro indicators
  - ../data/raw_data/interest_rates.csv      — Federal Funds rate (FRED)

OUTPUTS:
  - ../data/processed_data/df_model.csv      — model-ready dataset

SECTIONS:
  A. Target variable construction (crisis flags from 4 external sources +
     internal balance-sheet triggers)
  B. Feature engineering (ratios, rolling stats, compositions, growth rates)
  C. Macroeconomic variable integration
  D. Final model dataset assembly and export
================================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.path = "../results/figures/03_prep-"
)
```


## 1. Load Libraries

```{r load-libraries}
library(tidyverse)
library(lubridate)
library(skimr)
library(janitor)
library(haven)
library(readxl)
```

## 2. Load Cleaned Panel

```{r load-data}
df_clean <- readRDS("../data/processed_data/df_clean.rds")
```




# Part A: Target Variable Construction — Crisis Flags

The composite **`crisis`** target is built from multiple external data
sources (regulatory enforcement orders, bank failures) and internal
balance-sheet triggers (net loss, critically undercapitalised). Each source
is joined to the bank panel on the *previous* year (i.e., the flag at
year *t* indicates an event at year *t + 1*) so the model predicts future
deterioration.

## 3. FDIC Enforcement Orders (EDO)

Load FDIC Enforcement & Disciplinary Orders and match them to banks via
`id_fdic_cert`. The join uses `year == prev_year` so that a notice issued
in year *t+1* creates a flag on year-*t* features.

```{r prep-fdic-cert}
# Ensure FDIC certificate ID is numeric for joining
df_clean <- df_clean %>%
  mutate(id_fdic_cert = as.numeric(id_fdic_cert))
```
```{r load-notices}
# Read FDIC Enforcement & Disciplinary Orders
notices <- read_csv("../data/raw_data/EDOOrders.csv")
head(notices)
```

```{r parse-notice-dates}
# Parse the order's issued date and extract the year
notices <- notices %>%
  mutate(
    issued_date = as.Date(`Issued Date`),
    year = year(issued_date)
  )
```

```{r clean-notices}
# Select relevant columns and rename for consistency
notices_clean <- notices %>%
  select(
    id_fdic_cert = `Cert Number`,
    year,
    order_title = `Order Title`
  )
df_clean <- df_clean %>%
  mutate(id_fdic_cert = as.numeric(id_fdic_cert))

# Clean the certificate number (strip trailing semicolons) and
# create prev_year for lead-target join logic.
notices_clean <- notices_clean %>%
  mutate(id_fdic_cert = str_remove(id_fdic_cert, ";$"),
    id_fdic_cert = as.numeric(id_fdic_cert),
    prev_year = year - 1)
```

```{r join-notices}
# Join: match bank features at year t with notices issued at year t+1
df_clean <- df_clean %>%
  left_join(
    notices_clean,
    by = c("id_fdic_cert", "year" = "prev_year")
  )
```

```{r create-notice-flag}
# Binary flag: 1 if the bank received an FDIC regulatory notice, 0 otherwise
df_clean <- df_clean %>%
  mutate(
    notice_flag = ifelse(is.na(order_title), 0, 1)
  )
```

```{r count-notices}
# Sanity check: total banks with FDIC enforcement notices
sum(df_clean$notice_flag == 1, na.rm = TRUE)
```
```{r plot-notices}
# Visualise the temporal distribution of regulatory notices
df_clean %>%
  group_by(year) %>%
  summarise(notices = sum(notice_flag == 1, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = notices)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Number of Regulatory Notices per Year",
    x = "Year",
    y = "Number of Notices"
  ) +
  theme_minimal()
```


## 4. OCC Enforcement Actions — Historical ("Old")

Load historical OCC (Office of the Comptroller of the Currency) enforcement
orders. These target thrift institutions and use the Docket Number as the
join key (`id_thrift`).

```{r load-occ-old}
occ_old <- read_excel("../data/raw_data/occ_old.xlsx")
occ_old <- occ_old %>%
  rename(id_thrift = `Docket Number`)
```
```{r clean-occ-old}
# Parse issue date (handles multiple date formats), extract year,
# and set occ_old_order = 1 as an indicator
occ_old_clean <- occ_old %>%
  mutate(
    `Issue Date` = trimws(`Issue Date`),
    `Issue Date` = na_if(`Issue Date`, ""),
    issue_date = parse_date_time(
      `Issue Date`,
      orders = c("ymd HMS", "ymd", "dmy", "mdy")
    ),
    issue_date = as.Date(issue_date),
    year = year(issue_date),
    prev_year = year - 1,
    occ_old_order = 1
  ) %>%
  select(id_thrift, prev_year, occ_old_order)
```
```{r parse-occ-old-id}
# Strip the leading "H" from thrift docket numbers and convert to numeric
occ_old_clean <- occ_old_clean %>%
  mutate(
    id_thrift = str_remove(id_thrift, "^H"),
    id_thrift = as.numeric(id_thrift)
  )
```

```{r join-occ-old}
# Join old OCC orders to the bank panel (year t features ↔ year t+1 order)
df_clean <- df_clean %>%
  left_join(
    occ_old_clean,
    by = c("id_thrift", "year" = "prev_year")
  )
```

```{r create-occ-old-flag}
# Binary flag for old OCC enforcement action
df_clean <- df_clean %>%
  mutate(
    occ_old_flag = ifelse(is.na(occ_old_order), 0, 1)
  )
```

```{r count-occ-old}
# Sanity check: total banks with old OCC orders
sum(df_clean$occ_old_flag == 1, na.rm = TRUE)
```


## 5. OCC Enforcement Actions — Recent ("New")

Load the more recent OCC enforcement actions. These use `CharterNumber`
(renamed to `id_occ`) as the join key.

```{r load-occ-new}
occ_new <- read.csv("../data/raw_data/occ_new.csv")
occ_new <- occ_new %>%
  rename(id_occ = CharterNumber)
```

```{r clean-occ-new}
# Parse start date, extract year, create prev_year for lead-target join
occ_new_clean <- occ_new %>%
  mutate(
    issue_date = parse_date_time(
      StartDate,
      orders = c("dmy", "mdy", "ymd")
    ),
    issue_date = as.Date(issue_date),
    year = year(issue_date),
    prev_year = year - 1,
    occ_new_order = `Institution`
  ) %>%
  select(id_occ, prev_year, occ_new_order = `Institution`)
```

```{r join-occ-new}
# Join new OCC orders to bank panel
df_clean <- df_clean %>%
  left_join(
    occ_new_clean,
    by = c("id_occ", "year" = "prev_year")
  )
```

```{r occ-flag}
# Binary flag for new OCC enforcement action
df_clean <- df_clean %>%
  mutate(
    occ_new_flag = ifelse(is.na(occ_new_order), 0, 1)
  )
```

```{r check-occ}
# Sanity check: total banks with new OCC orders
sum(df_clean$occ_new_flag == 1, na.rm = TRUE)
```



## 6. FDIC Bank Failures

Load the FDIC bank failure list. A bank that fails in year *t+1*
gets a `fail_flag = 1` on its year-*t* record.

```{r load-failures}
failures <- read.csv("../data/raw_data/failures.csv")
failures <- failures %>%
  rename(id_fdic_cert = CERT)
```

```{r clean-failures}
# Parse the failure date, extract year, create prev_year for the lead join
failures_clean <- failures %>%
  mutate(
    FAILDATE = trimws(FAILDATE),
    FAILDATE = na_if(FAILDATE, ""),
    fail_date = parse_date_time(FAILDATE, orders = c("dmy", "mdy", "ymd")),
    fail_date = as.Date(fail_date),
    year = year(fail_date),
    prev_year = year-1,
    fail = 1
  )
```

```{r join-failures}
# Join failures to bank panel (year t features <-> year t+1 failure)
df_clean <- df_clean %>%
  left_join(
    failures_clean,
    by = c("id_fdic_cert", "year" = "prev_year")
  )
```

```{r create-fail-flag}
# Binary flag: 1 if the bank failed the following year
df_clean <- df_clean %>%
  mutate(
    fail_flag = ifelse(is.na(fail), 0, 1)
  )
```

```{r count-failures}
# Sanity check: total bank-years followed by a failure
sum(df_clean$fail_flag == 1, na.rm = TRUE)
```

```{r plot-failures}
# Visualise the temporal distribution of bank failures
df_clean %>%
  group_by(year) %>%
  summarise(fails = sum(fail_flag == 1, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = fails)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Number of Failures per Year",
    x = "Year",
    y = "Number of Failures"
  ) +
  theme_minimal()
```



## 7. Internal Balance-Sheet Targets

### 7.1 Critically Undercapitalised Flag

A bank is critically undercapitalised if its equity-to-assets ratio falls
below 2% (the PCA threshold). Again, `lead()` shifts to the next year.

```{r undercapitalized-flag}
df_clean <- df_clean %>%
  arrange(id_rssd, year) %>%
  group_by(id_rssd) %>%
  mutate(
    equity_ratio = ifelse(assets > 0, equity / assets, NA_real_),
    critically_undercapitalized_flag = lead(ifelse(equity_ratio < 0.02, 1, 0), 1)
  ) %>%
  ungroup()
```

```{r count-undercapitalized}
sum(df_clean$critically_undercapitalized_flag == 1, na.rm = TRUE)
```
`


## 8. Clean Up and Composite Crisis Target

### 8.1 Drop Auxiliary Columns

Remove raw text / metadata columns brought in from the external sources —
only the binary flags are needed going forward.

```{r drop-aux-cols}
df_clean <- df_clean %>%
  select(-c(
    order_title,
    occ_old_order,
    occ_new_order,
    CITYST,
    FAILDATE,
    COST,
    FIN,
    ID,
    NAME,
    QBFASSET,
    QBFDEP,
    RESTYPE,
    RESTYPE1,
    SAVR,
    year.y,
    fail))
```

### 8.2 Create Composite `crisis` Target

A bank-year is labelled as **crisis = 1** if *any* of the six individual
flags fire. This provides a single binary target for the downstream
classification models.

```{r create-crisis-target}
df_clean <- df_clean %>%
  mutate(
    crisis = if_else(
      if_any(c(notice_flag, occ_new_flag, occ_old_flag, fail_flag, critically_undercapitalized_flag), ~ .x == 1),
      1L, 0L
    )
  )
```





## 9. Diagnostic Visualisations — Target Distribution

```{r count-crisis}
# Total crisis bank-years in the dataset
sum(df_clean$crisis == 1, na.rm = TRUE)
```

### 9.1 Decomposition by Flag Component

```{r flag-breakdown}
p <- df_clean %>%
  group_by(year) %>%
  summarise(
    `FDIC Notice`       = sum(notice_flag == 1, na.rm = TRUE),
    `OCC Notice`        = sum(pmax(occ_old_flag, occ_new_flag, na.rm = TRUE) == 1, na.rm = TRUE),
    `Bank Failure`      = sum(fail_flag == 1, na.rm = TRUE),
    `Critically Undercapitalized` = sum(critically_undercapitalized_flag == 1, na.rm = TRUE),
    `Total Flags`       = sum(crisis == 1, na.rm = TRUE)
  ) %>%
  mutate(year = year - 1) %>%
  pivot_longer(-year, names_to = "Flag", values_to = "Count") %>%
  ggplot(aes(x = year, y = Count, color = Flag)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  scale_color_manual(values = c(
    "Bank Failure" = "#E41A1C",
    "Critically Undercapitalized" = "#377EB8",
    "FDIC Notice" = "#4DAF4A",
    "OCC Notice" = "#984EA3",
    "Total Flags" = "lightgrey"
  )) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "Year",
    y = "Number of Banks",
    color = "Flag Type", 
    title = "Flag Breakdown over Time"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom",
    
    # 🔽 increase sizes
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 11),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 11),
    
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

ggsave(
  filename = "../results/figures/flag_breakdown.png",
  plot = p,
  width = 8,
  height = 5,
  dpi = 300
)
```

```{r plot-p}
p
```
### 9.2 Composite Crisis Count Over Time

```{r plot-crisis-count}
df_clean %>%
  group_by(year) %>%
  summarise(n_crisis = sum(crisis == 1, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = n_crisis)) +
  geom_line(color = "#3266ad", linewidth = 0.8) +
  geom_point(color = "#3266ad", size = 2) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Number of crisis banks per year",
    x = "Year",
    y = "Number of banks"
  ) +
  theme_minimal()
```

### 9.3 Crisis vs Total Banks

```{r plot-crisis-vs-total}
# Overlay crisis banks (blue) against the total bank population (red)
df_clean %>%
  group_by(year) %>%
  summarise(
    n_crisis = sum(crisis == 1, na.rm = TRUE),
    n_banks = n()  # total banks per year
  ) %>%
  ggplot(aes(x = year)) +
  geom_line(aes(y = n_crisis), color = "#3266ad", linewidth = 0.8) +
  geom_line(aes(y = n_banks), color = "#e63946", linewidth = 0.8) +
  labs(
    title = "Number of Crisis Banks and Total Banks per Year",
    x = "Year",
    y = "Number of Banks",
    caption = "Blue: crisis banks; Red: total banks"
  ) +
  theme_minimal()
```

### 9.4 Crisis Rate (post-1980)

```{r crisis-rate}
# Count and rate of crisis observations from 1980 onwards
sum(df_clean$crisis == 1 & df_clean$year >= 1980, na.rm = TRUE)
sum(df_clean$year >= 1980, na.rm = TRUE)
sum(df_clean$crisis == 1 & df_clean$year >= 1980, na.rm = TRUE)/sum(df_clean$year >= 1980, na.rm = TRUE)
```

### 9.5 Filter to Post-1980

Restrict the panel to years >= 1980 to align with macroeconomic data
availability and regulatory regime relevance.

```{r filter-post-1978}
df_clean <- df_clean %>%
  filter(year >= 1980)
```




# Part B: Macroeconomic Control Variables

## 10. World Bank Macro Indicators

Load annual U.S. macroeconomic series from the World Bank and join them
to the bank panel by year. These provide system-wide context that helps
models distinguish idiosyncratic bank distress from macro-driven stress.

```{r load-macro}
macro <- read.csv("../data/raw_data/macrodata.csv", stringsAsFactors = FALSE)
head(macro)
```

```{r clean-macro}
macro_clean <- macro %>%
  # Drop metadata columns (country identifiers, time code)
  select(-Country.Name, -Country.Code, -Time.Code) %>%
  # Rename Time to year
  rename(year = Time) %>%
  # Rename World Bank indicator codes to human-readable snake_case
  rename(
    pop_growth     = Population.growth..annual.....SP.POP.GROW.,
    pop_total      = Population..total..SP.POP.TOTL.,
    gdp_per_capita = GDP.per.capita..current.US....NY.GDP.PCAP.CD.,
    gdp            = GDP..current.US....NY.GDP.MKTP.CD.,
    gdp_growth     = GDP.growth..annual.....NY.GDP.MKTP.KD.ZG.,
    gdp_pc_growth  = GDP.per.capita.growth..annual.....NY.GDP.PCAP.KD.ZG.,
    real_int_rate  = Real.interest.rate......FR.INR.RINR.,
    inflation      = Inflation..consumer.prices..annual.....FP.CPI.TOTL.ZG.,
    unemployment   = Unemployment..total....of.total.labor.force...national.estimate...SL.UEM.TOTL.NE.ZS.
  ) %>%
  # Replace World Bank missing-value marker ".." with NA, then convert to numeric
  mutate(across(-year, ~ as.numeric(na_if(., "..")))) %>%
  # Drop footer rows where year is NA
  filter(!is.na(year))
```

```{r join-macro}
# Merge macro indicators into the bank panel on year
df_clean <- df_clean %>%
  left_join(macro_clean, by = "year")
```

```{r check-macro}
names(macro_clean)
summary(macro_clean)
```

## 11. Federal Funds Rate (FRED)

Load monthly Federal Funds effective rate from FRED, compute an annual
average, and merge it into the bank panel.

```{r load-rates}
rates <- read.csv("../data/raw_data/interest_rates.csv", stringsAsFactors = FALSE)
head(rates)
```

```{r clean-rates}
# Convert to annual average (monthly observations -> one value per year)
rates_clean <- rates %>%
  mutate(
    observation_date = as.Date(observation_date),
    year = year(observation_date)
  ) %>%
  rename(fed_funds_rate = FEDFUNDS) %>%
  group_by(year) %>%
  summarise(fed_funds_rate = mean(fed_funds_rate, na.rm = TRUE)) %>%
  ungroup()
```

```{r join-rates}
df_clean <- df_clean %>%
  left_join(rates_clean, by = "year")
```

```{r check-rates}
summary(rates_clean)
```



# Part C: Feature Engineering

## 12. Financial Ratios

Compute six core financial ratios that capture capital adequacy,
profitability, asset quality, and funding cost.

```{r financial-ratios}
df_clean <- df_clean %>%
  mutate(
    # Capital Adequacy: Total Equity / Total Assets
    capital_ratio = ifelse(assets > 0, equity / assets, NA_real_),

    # Profitability: Return on Assets = Net Income / Total Assets
    roa = ifelse(assets > 0, ytdnetinc / assets, NA_real_),

    # Profitability: Return on Equity = Net Income / Total Equity
    roe = ifelse(equity > 0, ytdnetinc / equity, NA_real_),

    # Asset Quality: Non-Performing Loans / Total Loans
    npl_ratio = ifelse(ln_tot_gross > 0, npl_tot / ln_tot_gross, NA_real_),

    # Net Interest Margin: (Interest Income - Interest Expense) / Total Assets
    nim = ifelse(assets > 0, (ytdint_inc - ytdint_exp)/ assets, NA_real_),

    # Cost of Funding: Total Interest Expense / (Deposits + Other Borrowings)
    cost_of_funding = ifelse(
      (deposits + othbor_liab) > 0,
      ytdint_exp / (deposits + othbor_liab),
      NA_real_
    )
  )
```

```{r income-growth-flag}
df_clean <- df_clean %>%
  group_by(id_rssd) %>%
  mutate(
    growth = ifelse(
      !is.na(lead(ytdnetinc, 1)) & !is.na(ytdnetinc) &
      ytdnetinc > 0 &                                        # currently profitable
      lead(ytdnetinc, 1) > 0 &                               # profitable next year
      lead(ytdnetinc, 1) >= ytdnetinc * 1.30 &               # 30%+ income growth
      lead(roe) - roe > 0.05,                                # 5 percentage point roe increase
      1L, 0L
    )
  ) %>%
  ungroup()
```
## 13. Rolling 3-Year Statistics

Compute the standard deviation of key variables over a 3-year window
(years *t*, *t-1*, *t-2*) for each bank. Higher rolling volatility may
signal instability preceding a crisis.

```{r rolling-stats}
df_clean <- df_clean %>%
  arrange(id_rssd, year) %>%
  group_by(id_rssd) %>%
  mutate(
    #  Equity 
    equity_sd_3y   = pmap_dbl(
      list(equity, lag(equity, 1), lag(equity, 2)),
      ~ sd(c(..1, ..2, ..3), na.rm = FALSE)
    ),

    #  Assets 
    assets_sd_3y   = pmap_dbl(
      list(assets, lag(assets, 1), lag(assets, 2)),
      ~ sd(c(..1, ..2, ..3), na.rm = FALSE)
    ),

    #  Total Loans 
    loans_sd_3y   = pmap_dbl(
      list(ln_tot_gross, lag(ln_tot_gross, 1), lag(ln_tot_gross, 2)),
      ~ sd(c(..1, ..2, ..3), na.rm = FALSE)
    ),

    #  Deposits 
    deposits_sd_3y   = pmap_dbl(
      list(deposits, lag(deposits, 1), lag(deposits, 2)),
      ~ sd(c(..1, ..2, ..3), na.rm = FALSE)
    ),

    #  Net Income (earnings volatility proxy) 
    netinc_sd_3y   = pmap_dbl(
      list(ytdnetinc, lag(ytdnetinc, 1), lag(ytdnetinc, 2)),
      ~ sd(c(..1, ..2, ..3), na.rm = FALSE)
    ),

    #  ROA (return on assets) 
    roa_sd_3y   = pmap_dbl(
      list(roa, lag(roa, 1), lag(roa, 2)),
      ~ sd(c(..1, ..2, ..3), na.rm = FALSE)
    )
  ) %>%
  ungroup()
```

## 14. Loan Composition

Fraction of total gross loans allocated to each major loan category.
Concentration in a single loan type (e.g., real estate) is a known risk
factor.

```{r loan-proportions}
df_clean <- df_clean %>%
  mutate(
    prop_ln_re   = ifelse(ln_tot_gross > 0, ln_re / ln_tot_gross, NA_real_),   # Real estate
    prop_ln_ci   = ifelse(ln_tot_gross > 0, ln_ci / ln_tot_gross, NA_real_),   # Commercial & industrial
    prop_ln_cons = ifelse(ln_tot_gross > 0, ln_cons / ln_tot_gross, NA_real_), # Consumer
    prop_ln_agr  = ifelse(ln_tot_gross > 0, ln_agr / ln_tot_gross, NA_real_),  # Agriculture
    prop_ln_cc = ifelse(ln_tot_gross > 0, ln_cc / ln_tot_gross, NA_real_),  # Credit cards
    prop_ln_fi = ifelse(ln_tot_gross > 0, ln_fi / ln_tot_gross, NA_real_)  # Financial Institutions
  )
```

## 15. Deposit Composition

Proportion of demand (non-interest-bearing) deposits in total deposits.
Banks heavily reliant on volatile demand deposits may face liquidity risk.

```{r deposit-proportions}
df_clean <- df_clean %>%
  mutate(
    prop_demand_dep = ifelse(deposits > 0, demand_deposits / deposits, NA_real_),
  )
```

## 16. Year-over-Year Growth Rates and Ratio Changes

Captures the direction and magnitude of balance-sheet expansion or
contraction. Rapid asset/loan growth may precede credit-quality problems;
declining ratios may signal deterioration.

```{r growth-rates}
df_clean <- df_clean %>%
  arrange(id_rssd, year) %>%
  group_by(id_rssd) %>%
  mutate(
    # Percentage growth rates (YoY)
    asset_growth    = (assets       - lag(assets))       / lag(assets),
    deposit_growth  = (deposits     - lag(deposits))     / lag(deposits),
    loan_growth     = (ln_tot_gross - lag(ln_tot_gross)) / lag(ln_tot_gross),
    liab_growth     = (liab_tot     - lag(liab_tot))     / lag(liab_tot),
    equity_growth   = (equity       - lag(equity))       / lag(equity),

    # Year-over-year changes in financial ratios (percentage-point changes)
    capital_ratio_chg   = capital_ratio   - lag(capital_ratio),
    roa_chg             = roa             - lag(roa),
    roe_chg             = roe             - lag(roe),
    nim_chg             = nim             - lag(nim),
    cost_of_funding_chg = cost_of_funding - lag(cost_of_funding),
    npl_chg = npl_ratio - lag(npl_ratio)
  ) %>%
  ungroup()
```

## 17. Bank Age

```{r bank-age}
# Bank age: number of years since the bank's charter/opening date
df_clean <- df_clean %>%
  mutate(age = year - open_year)
```

## 18. Outlier Capping (Winsorisation)

Cap extreme ratio and growth-rate values that are likely data errors or
non-representative edge cases. Values outside plausible bounds are set
to `NA`.

```{r cap-outliers}
# IQR-based outlier removal function
remove_outliers_iqr <- function(x, k = 3) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - k * iqr
  upper <- q3 + k * iqr
  ifelse(x < lower | x > upper, NA, x)
}

outlier_cols <- c(
  # Financial ratios
  "roa", "roe", "capital_ratio", "npl_ratio", "nim", "cost_of_funding",
  
  # Growth rates
  "asset_growth", "loan_growth", "deposit_growth", "liab_growth", "equity_growth",
  
  # Ratio changes
  "capital_ratio_chg", "roa_chg", "roe_chg", "nim_chg", "cost_of_funding_chg", "npl_chg",
  
  # Rolling 3-year std devs
  "equity_sd_3y", "assets_sd_3y", "loans_sd_3y", "deposits_sd_3y",
  "netinc_sd_3y", "roa_sd_3y",
  
  # Intra-year dynamics
  "assets_sd_intra", "loans_sd_intra", "deposits_sd_intra", "equity_sd_intra",
  "assets_max_decline", "loans_max_decline", "deposits_max_decline", "equity_max_decline",
  
  #Size
  "num_employees"
)

df_clean <- df_clean %>%
  mutate(across(all_of(outlier_cols), ~remove_outliers_iqr(., k = 3)))


```

# Part D: Final Model Dataset Assembly

## 19. Select Model Features

Assemble the final feature matrix by selecting only the variables needed
for modelling. Features are grouped into logical categories:

- **Identifiers** — kept for data splitting and reference, dropped before training
- **Targets** — `crisis` (binary composite), `growth` (income growth flag)
- **Balance-sheet levels** — size, capital, asset quality proxies
- **Income statement** — non-interest income/expense, staffing, age
- **Financial ratios** — capital adequacy, profitability, asset quality, NIM, funding cost
- **Growth rates** — YoY changes in major balance-sheet items
- **Ratio changes** — YoY percentage-point shifts in key ratios
- **Rolling stats** — 3-year standard deviations capturing medium-term volatility
- **Composition** — loan-type and deposit-type proportions
- **Intra-year dynamics** — within-year (quarterly) volatility and max declines
- **Macroeconomic** — GDP growth, inflation, unemployment, interest rates

```{r create-model-dataset}
df_model <- df_clean %>%
  select(
    #  Identifiers (keep for splitting / reference, drop before training) 
    id_rssd, year,

    #  Target 
    crisis, growth,

    #  Balance sheet levels 
    assets, equity, ln_tot_gross, deposits, liab_tot,
    cash, securities, llres, fixed_ass, oreo, othbor_liab, subdebt,

    #  Income statement & size 
    ytdnonint_inc, ytdnonint_exp, ytdllprov, ytdnetinc, ytdint_inc, ytdint_exp,
    num_employees, age,

    #  Financial ratios 
    capital_ratio, roa, roe, nim, cost_of_funding, npl_ratio,

    #  Growth rates 
    asset_growth, deposit_growth, loan_growth, liab_growth, equity_growth,

    #  Ratio changes (pp) 
    capital_ratio_chg, roa_chg, roe_chg, nim_chg, cost_of_funding_chg, npl_chg,

    #  Rolling 3-year stats 
    equity_sd_3y, assets_sd_3y,
    loans_sd_3y, deposits_sd_3y,
    netinc_sd_3y, roa_sd_3y,

    #  Loan composition 
    prop_ln_re, prop_ln_ci, prop_ln_cons, prop_ln_agr, prop_ln_cc, prop_ln_fi, 

    #  Deposit composition 
    prop_demand_dep,

    #  Intra-year dynamics (quarterly seasonality features) 
    assets_sd_intra, loans_sd_intra, deposits_sd_intra, equity_sd_intra,
    assets_max_decline, loans_max_decline, deposits_max_decline, equity_max_decline,

    #  Macroeconomic 
    gdp, inflation, unemployment, real_int_rate,
    pop_total, fed_funds_rate,
  ) %>%
  # Drop rows with no target or outside the study period
  filter(!is.na(crisis) & year >= 1980 & year <= 2023 & !is.na(growth))

cat("Model dataset dimensions:", nrow(df_model), "rows x", ncol(df_model), "columns\n")
cat("Crisis rate:", round(mean(df_model$crisis, na.rm = TRUE) * 100, 2), "%\n")
cat("Growth rate:", round(mean(df_model$growth, na.rm = TRUE) * 100, 2), "%\n")
```



## 20. Export Model-Ready Dataset

```{r save-model-data}
write.csv(df_model, "../data/processed_data/df_model.csv", row.names = FALSE)
cat("Saved df_model.csv\n")
```
