---
title: "Feature–Target Exploratory Analysis"
output: pdf_document
date: "2026-04-24"
---

<!--
================================================================================
04_feature_target_eda.Rmd

PURPOSE:
  Investigate the relationship between every feature in `df_model` and the
  binary target `crisis`. This notebook is the justification layer for the
  feature set used in downstream modelling: for every feature we show (a)
  its univariate distribution and (b) how that distribution differs between
  crisis and non-crisis bank-years. The goal is to demonstrate empirically
  that the chosen features carry signal about future distress.

WORKFLOW POSITION:
  1. 01_data_exploration.Rmd
  2. 02_data_cleaning.Rmd
  3. 03_data_preparation.Rmd
  4. **04_feature_target_eda.Rmd** ← (this file)
  5. Model training scripts (XGBoost, LightGBM, Random Forest, etc.)

INPUTS:
  - ../data/processed_data/df_model.csv  — model-ready dataset

OUTPUTS:
  - Inline figures and tables only (no artefacts written to disk)

SECTIONS:
  A. Statistical discrimination — Wilcoxon tests and standardised
     mean differences (Cohen's d) to rank features by signal strength
  B. Bivariate analysis — feature distribution split by crisis status
  C. Group-wise summary statistics (mean / median / SD by crisis)
  D. Stratified subgroup analysis — crisis rates and Cohen's d of top
     features within bins of age, employee count, and asset size
  E. Pre-crisis feature trajectories — median drift of key features in
     the five years preceding crisis events, benchmarked to the
     non-crisis baseline
================================================================================
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  dev      = "png",
  fig.width  = 9,
  fig.height = 6,
  dpi = 300,
  fig.path = "../results/figures/04_eda-"
)
```

## 1. Load Libraries

```{r load-libraries}
library(tidyverse)
library(scales)
library(knitr)
library(effsize)
library(car)
library(gridExtra)
```

## 2. Load Model-Ready Data

```{r load-data}
df_model <- read_csv("../data/processed_data/df_model.csv",
                     show_col_types = FALSE)

cat("df_model dimensions:", nrow(df_model), "rows x", ncol(df_model), "columns\n")
cat("Crisis rate:", round(mean(df_model$crisis, na.rm = TRUE) * 100, 2), "%\n")

df_model <- df_model %>%
  mutate(crisis_lbl = factor(crisis,
                             levels = c(0, 1),
                             labels = c("Non-crisis", "Crisis")))
```

---

## 3. Design System

A single theme and two-colour palette are used across every chart in this
notebook so visuals stay comparable. Blue denotes non-crisis bank-years,
red denotes crisis bank-years — the same convention used throughout the
project.

```{r design-system}
# --- Palette ---------------------------------------------------------------
crisis_colors <- c("Non-crisis" = "#3266ad",   # project blue
                   "Crisis"     = "#e63946")   # project red

# --- Theme -----------------------------------------------------------------
eda_theme <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title        = element_text(face = "bold", size = base_size + 1),
      plot.subtitle     = element_text(colour = "grey35", size = base_size - 1),
      plot.caption      = element_text(colour = "grey45", size = base_size - 2),
      panel.grid.minor  = element_blank(),
      panel.grid.major  = element_line(colour = "grey92"),
      axis.line         = element_line(colour = "grey70"),
      strip.text        = element_text(face = "bold", size = base_size - 1),
      legend.position   = "top",
      legend.title      = element_blank(),
      legend.key.height = unit(0.4, "cm")
    )
}


```

---

## 4. Feature Groups

Features are organised into the same thematic blocks used in
`03_data_preparation.Rmd`. Grouping keeps the univariate and bivariate
panels readable and lets us interpret each block (e.g., capital adequacy,
asset quality) as a whole.

```{r feature-groups}
feature_groups <- list(
  "Balance sheet (log)" = c("assets", "equity", "ln_tot_gross", "deposits",
                            "liab_tot", "cash", "securities", "llres",
                            "fixed_ass", "oreo", "othbor_liab", "subdebt"),
  "Income & staffing"   = c("ytdnonint_inc", "ytdnonint_exp",
                            "num_employees", "age"),
  "Financial ratios"    = c("capital_ratio", "roa", "roe", "nim",
                            "cost_of_funding", "npl_ratio"),
  "Growth rates"        = c("asset_growth", "deposit_growth", "loan_growth",
                            "liab_growth", "equity_growth"),
  "Ratio changes"       = c("capital_ratio_chg", "roa_chg", "roe_chg",
                            "nim_chg", "cost_of_funding_chg", "npl_chg"),
  "Rolling 3y std dev"  = c("equity_sd_3y", "assets_sd_3y",
                            "loans_sd_3y", "deposits_sd_3y",
                            "netinc_sd_3y", "roa_sd_3y"),
  "Composition"         = c("prop_ln_re", "prop_ln_ci", "prop_ln_cons",
                            "prop_ln_agr", "prop_ln_cc", "prop_ln_fi", "prop_demand_dep"),
  "Intra-year dynamics" = c("assets_sd_intra", "loans_sd_intra",
                            "deposits_sd_intra", "equity_sd_intra",
                            "assets_max_decline", "loans_max_decline",
                            "deposits_max_decline", "equity_max_decline"),
  "Macroeconomic"       = c("gdp", "inflation", "unemployment",
                            "real_int_rate", "pop_total", "fed_funds_rate")
)

# Flatten to a (feature, group) lookup
feat_lookup <- feature_groups %>%
  enframe(name = "group", value = "feature") %>%
  unnest(feature)

feature_names <- c(
  "assets" = "Assets",
  "equity" = "Equity",
  "deposits" = "Deposits",
  "ln_tot_gross" = "Loans",
  "liab_tot" = "Liabilities",
  "ytdnetinc" = "Net Income",
  "cash" = "Cash",
  "securities" = "Securities",
  "llres" = "Loan Loss Reserves",
  "fixed_ass" = "Fixed Assets",
  "oreo" = "Other Real Estate Owned",
  "othbor_liab" = "Other Borrowed Liabilities",
  "subdebt" = "Subordinated Debt",
  "ytdnonint_inc" = "Non-interest Income",
  "ytdnonint_exp" = "Non-interest Expense",
  "num_employees" = "Number of Employees",
  "age" = "Bank Age",
  "capital_ratio" = "Capital Ratio",
  "roa" = "ROA",
  "roe" = "ROE",
  "nim" = "NIM",
  "cost_of_funding" = "Cost of Funding",
  "npl_ratio" = "NPL Ratio",
  "asset_growth" = "Asset Growth",
  "deposit_growth" = "Deposit Growth",
  "loan_growth" = "Loan Growth",
  "liab_growth" = "Liability Growth",
  "equity_growth" = "Equity Growth",
  "capital_ratio_chg" = "Capital Ratio Change",
  "roa_chg" = "ROA Change",
  "roe_chg" = "ROE Change",
  "nim_chg" = "NIM Change",
  "cost_of_funding_chg" = "Cost of Funding Change",
  "npl_chg" = "NPL Change",
  "equity_sd_3y" = "Equity Volatility (3Y)",
  "assets_sd_3y" = "Assets Volatility (3Y)",
  "loans_sd_3y" = "Loans Volatility (3Y)",
  "deposits_sd_3y" = "Deposits Volatility (3Y)",
  "netinc_sd_3y" = "Net Income Volatility (3Y)",
  "roa_sd_3y" = "ROA Volatility (3Y)",
  "prop_ln_re" = "Prop. Real Estate Loans",
  "prop_ln_ci" = "Prop. C&I Loans",
  "prop_ln_cons" = "Prop. Consumer Loans",
  "prop_ln_agr" = "Prop. Agricultural Loans",
  "prop_ln_cc" = "Prop. Credit Cards",
  "prop_ln_fi" = "Prop. Financial Institutions",
  "prop_demand_dep" = "Prop. Demand Deposits",
  "assets_sd_intra" = "Assets Volatility (Intra)",
  "loans_sd_intra" = "Loans Volatility (Intra)",
  "deposits_sd_intra" = "Deposits Volatility (Intra)",
  "equity_sd_intra" = "Equity Volatility (Intra)",
  "assets_max_decline" = "Assets Max Decline",
  "loans_max_decline" = "Loans Max Decline",
  "deposits_max_decline" = "Deposits Max Decline",
  "equity_max_decline" = "Equity Max Decline",
  "gdp" = "GDP",
  "inflation" = "Inflation",
  "unemployment" = "Unemployment Rate",
  "real_int_rate" = "Real Interest Rate",
  "pop_total" = "Population Total",
  "fed_funds_rate" = "Fed Funds Rate"
)

```

Features that are strictly positive and heavily right-skewed
(balance-sheet levels) are log-transformed for visualisation only. All
statistical tests use the raw values.

```{r log-set}
log_features <- feature_groups[["Balance sheet (log)"]]
```


# Part A — Statistical Discrimination

Distributional differences are quantified with two complementary measures:

1. **Wilcoxon rank-sum test** — non-parametric, no normality assumption.
   Reports whether the two conditional distributions differ at all.
2. **Cohen's d** — standardised mean difference. Interpretable effect
   size; |d| > 0.2 = small, > 0.5 = medium, > 0.8 = large.

A feature is a strong candidate predictor when the Wilcoxon p-value is
small **and** |d| is not negligible.


## 5. Feature Ranking by Effect Size

```{r discrim-stats}
compute_discrim <- function(feat) {
  x <- df_model[[feat]][df_model$crisis == 0]
  y <- df_model[[feat]][df_model$crisis == 1]

  # Drop NA *and* Inf / -Inf (growth rates can blow up when lag() is ~0)
  x <- x[is.finite(x)]
  y <- y[is.finite(y)]

  sdx <- if (length(x) > 1) sd(x) else NA_real_
  sdy <- if (length(y) > 1) sd(y) else NA_real_

  bad <- length(x) < 30 || length(y) < 30 ||
         is.na(sdx) || is.na(sdy) ||
         sdx == 0 || sdy == 0

  if (bad) {
    return(tibble(feature = feat,
                  p_wilcox = NA_real_,
                  cohen_d  = NA_real_))
  }

  w <- suppressWarnings(wilcox.test(x, y, exact = FALSE))
  d <- tryCatch(effsize::cohen.d(y, x, na.rm = TRUE)$estimate,
                error = function(e) NA_real_)

  tibble(feature = feat,
         p_wilcox = w$p.value,
         cohen_d  = d)
}

discrim <- map_dfr(feat_lookup$feature, compute_discrim) %>%
  left_join(feat_lookup, by = "feature") %>%
  mutate(abs_d = abs(cohen_d)) %>%
  arrange(desc(abs_d))
```

```{r discrim-plot, fig.height=10}
discrim %>%
  filter(!is.na(cohen_d)) %>%
  mutate(feature_label = feature_names[feature]) %>%
  mutate(feature_label = fct_reorder(feature_label, abs_d),
         sign    = ifelse(cohen_d >= 0, "Higher in crisis", "Lower in crisis")) %>%
  ggplot(aes(x = cohen_d, y = feature_label, fill = sign)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, colour = "grey40", linewidth = 0.3) +
  geom_vline(xintercept = c(-0.2, 0.2),
             linetype = "dashed", colour = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = c("Higher in crisis" = crisis_colors["Crisis"],
                               "Lower in crisis"  = crisis_colors["Non-crisis"])) +
  labs(title = "Feature discrimination ranked by Cohen's d",
       subtitle = "Dashed lines = small-effect threshold (|d| = 0.2)",
       x = "Cohen's d (crisis - non-crisis, standardised)",
       y = NULL) +
  eda_theme()
ggsave("../results/figures/04_feature_discrimination_ranking.png", width = 10, height = 12)
```

## 6. Full Discrimination Table

```{r discrim-table}
discrim %>%
  transmute(
    Group   = group,
    Feature = feature_names[feature],
    `Cohen's d` = round(cohen_d, 3),
    `|d|`       = round(abs_d, 3),
    `Wilcoxon p` = signif(p_wilcox, 3),
    Direction = case_when(
      is.na(cohen_d)      ~ "—",
      cohen_d >=  0.2     ~ "Crisis ↑ (meaningful)",
      cohen_d <= -0.2     ~ "Crisis ↓ (meaningful)",
      TRUE                ~ "Small / negligible"
    )
  ) %>%
  kable(caption = "Wilcoxon rank-sum test and Cohen's d for every feature. Sorted by |d|.")
```

## 7. Share of Features with a Meaningful Effect

```{r discrim-summary}
discrim %>%
  filter(!is.na(cohen_d)) %>%
  mutate(bucket = case_when(
    abs_d <  0.2 ~ "Negligible (|d| < 0.2)",
    abs_d <  0.5 ~ "Small (0.2–0.5)",
    abs_d <  0.8 ~ "Medium (0.5–0.8)",
    TRUE         ~ "Large (> 0.8)"
  ),
  bucket = factor(bucket,
                  levels = c("Negligible (|d| < 0.2)",
                             "Small (0.2–0.5)",
                             "Medium (0.5–0.8)",
                             "Large (> 0.8)"))) %>%
  count(bucket) %>%
  mutate(share = n / sum(n)) %>%
  ggplot(aes(x = bucket, y = n, fill = bucket)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(n, "\n(", percent(share, 0.1), ")")),
            vjust = -0.2, size = 3.4, lineheight = 0.95) +
  scale_fill_manual(values = c("#c9d5e6", "#7fa0c7",
                               crisis_colors["Non-crisis"],
                               crisis_colors["Crisis"])) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  coord_cartesian(clip = "off") +
  labs(title = "How many features carry a meaningful crisis signal?",
       subtitle = "Buckets based on |Cohen's d|",
       x = NULL, y = "Number of features") +
  eda_theme() +
  theme(legend.position = "none")
```



# Part B — Bivariate Analysis: Feature vs Crisis

For every feature we now overlay the conditional distribution under
`crisis = 0` (blue) and `crisis = 1` (red). Visual separation between the
two distributions is direct evidence that the feature discriminates.

## 8. Density Plots by Crisis Status

```{r density-fn}
plot_density <- function(feats, title, log_x = FALSE) {
  sig_feats <- discrim %>% filter(feature %in% feats, abs_d >= 0.2) %>% pull(feature)
  
  d <- df_model %>%
    select(crisis_lbl, all_of(feats)) %>%
    pivot_longer(-crisis_lbl, names_to = "feature", values_to = "value") %>%
    filter(!is.na(value) & is.finite(value)) %>%
    mutate(feature_label = factor(feature_names[feature], levels = feature_names[feats]))
    
  sig_df <- tibble(
    feature_label = factor(feature_names[sig_feats], levels = feature_names[feats])
  )
  
  if (log_x) {
    d <- d %>% 
      filter(value > 0) %>%
      mutate(value = log10(value))
    xlab <- "log10(value)"
  } else {
    xlab <- "Value"
  }
  
  p <- ggplot(d, aes(x = value, fill = crisis_lbl, colour = crisis_lbl)) +
    geom_blank()
  
  if (nrow(sig_df) > 0) {
    p <- p + geom_rect(data = sig_df, aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf), 
                       fill = "ivory", inherit.aes = FALSE)
  }
  
p <- p +
  geom_density(alpha = 0.35, linewidth = 0.4) +
  facet_wrap(~ feature_label, scales = "free", ncol = 3) +
  scale_fill_manual(values = crisis_colors) +
  scale_colour_manual(values = crisis_colors) +
  labs(title = title,
       subtitle = "Ivory background indicates |Cohen's d| >= 0.2",
       x = xlab, y = "Density") +
  eda_theme() +
  theme(
    panel.background = element_rect(fill = "transparent", colour = NA),

    axis.title = element_text(size = 14),
    axis.text = element_text(size = 11),
    strip.text = element_text(size = 12),     
    legend.text = element_text(size = 11),
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),  # centered
    plot.subtitle = element_text(size = 12, hjust = 0.5, face = "italic")
  )
    
  return(p)
}
```

### 8.1 Balance-sheet levels

```{r den-balance, fig.height=8}
plot_density(log_features,
             "Balance Sheet Variables: Density by Crisis Status",
             log_x = TRUE)
```

### 8.2 Financial ratios

```{r den-ratios, fig.height=5}
plot_density(feature_groups[["Financial ratios"]],
             "Financial Ratios: Density by Crisis Status")
```

### 8.3 Growth Rates & Ratio Changes

```{r den-growth-ratio, fig.height=10}
plot_density(c(feature_groups[["Growth rates"]], feature_groups[["Ratio changes"]]),
             "YOY Growth Rates & Ratio Changes: Density by Crisis Status")
```

### 8.4 Rolling 3-year std dev

```{r den-rolling, fig.height=5}
plot_density(feature_groups[["Rolling 3y std dev"]],
             "Rolling volatility — density by crisis status")
```

### 8.5 Composition

```{r den-comp, fig.height=5}
plot_density(feature_groups[["Composition"]],
             "Composition — density by crisis status")
```

### 8.6 Loan Portfolio Composition — Crisis vs Non-Crisis

```{r pie-loan-composition, fig.height=5, fig.width=10}
# --- Loan-type proportion columns ---
loan_props <- c("prop_ln_re", "prop_ln_ci", "prop_ln_cons",
                "prop_ln_agr", "prop_ln_cc", "prop_ln_fi")

loan_labels <- c("Real Estate", "C&I", "Consumer",
                 "Agricultural", "Credit Cards", "Fin. Institutions")

# Average proportions per group
loan_avg <- df_model %>%
  group_by(crisis_lbl) %>%
  summarise(across(all_of(loan_props), ~ mean(.x, na.rm = TRUE)),
            .groups = "drop") %>%
  pivot_longer(-crisis_lbl, names_to = "loan_type", values_to = "proportion")

# Add "Other Loans" = 1 - sum of named types
other_loans <- loan_avg %>%
  group_by(crisis_lbl) %>%
  summarise(proportion = 1 - sum(proportion), .groups = "drop") %>%
  mutate(loan_type = "other")

loan_pie <- bind_rows(loan_avg, other_loans) %>%
  mutate(label = dplyr::recode(loan_type,
                        prop_ln_re   = "Real Estate",
                        prop_ln_ci   = "C&I",
                        prop_ln_cons = "Consumer",
                        prop_ln_agr  = "Agricultural",
                        prop_ln_cc   = "Credit Cards",
                        prop_ln_fi   = "Fin. Institutions",
                        other        = "Other Loans"))

# Colours — RColorBrewer Set1
loan_fill <- setNames(RColorBrewer::brewer.pal(7, "Set1"),
                      c("Real Estate", "C&I", "Consumer",
                        "Agricultural", "Credit Cards",
                        "Fin. Institutions", "Other Loans"))

# Build pie via coord_polar on a bar chart
ggplot(loan_pie,
       aes(x = "", y = proportion, fill = label)) +
  geom_col(width = 1, colour = "white", linewidth = 0.4) +
  geom_text(aes(label = ifelse(proportion >= 0.1,
                                paste0(round(proportion * 100, 1), "%"), "")),
            position = position_stack(vjust = 0.5),
            size = 4, colour = "grey15") +
  coord_polar(theta = "y") +
  facet_wrap(~ crisis_lbl) +
  scale_fill_manual(values = loan_fill) +
  labs(title = "Loan Portfolio Composition by Crisis Status",
       fill = NULL) +
  eda_theme() +
  theme(axis.title = element_blank(),
        axis.text  = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        strip.text = element_text(size = 12),      # facet title
        legend.text = element_text(size = 11),
        axis.line = element_blank(),
        plot.title = element_text(size = 15, face = "bold", hjust = 0.5))
```

### 8.7 Deposit Composition — Crisis vs Non-Crisis

```{r pie-deposit-composition, fig.height=5, fig.width=10}
# Demand deposits proportion is prop_demand_dep; time = 1 - demand
dep_avg <- df_model %>%
  group_by(crisis_lbl) %>%
  summarise(Demand = mean(prop_demand_dep, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(Time = 1 - Demand) %>%
  pivot_longer(-crisis_lbl, names_to = "deposit_type", values_to = "proportion")

dep_fill <- setNames(RColorBrewer::brewer.pal(3, "Set1")[1:2],
                     c("Demand", "Time"))

ggplot(dep_avg,
       aes(x = "", y = proportion, fill = deposit_type)) +
  geom_col(width = 1, colour = "white", linewidth = 0.4) +
  geom_text(aes(label = paste0(round(proportion * 100, 1), "%")),
            position = position_stack(vjust = 0.5),
            size = 4, colour = "grey15") +
  coord_polar(theta = "y") +
  facet_wrap(~ crisis_lbl) +
  scale_fill_manual(values = dep_fill) +
  labs(title = "Deposit Composition",
       subtitle = "Average share of demand vs time deposits — Crisis vs Non-crisis banks",
       fill = NULL) +
  eda_theme() +
  theme(axis.title = element_blank(),
        axis.text  = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

### 8.8 Intra-year dynamics

```{r den-intra, fig.height=7}
plot_density(feature_groups[["Intra-year dynamics"]],
             "Intra-year dynamics — density by crisis status")
```

### 8.9 Macroeconomic

```{r den-macro, fig.height=5}
plot_density(feature_groups[["Macroeconomic"]],
             "Macro environment — density by crisis status")
```

## 9. Boxplots by Crisis Status

Boxplots provide a complementary view: medians, IQRs, and tails per class.

```{r box-fn}
plot_box <- function(feats, title, log_y = FALSE) {
  sig_feats <- discrim %>% filter(feature %in% feats, abs_d >= 0.2) %>% pull(feature)
  
  d <- df_model %>%
    select(crisis_lbl, all_of(feats)) %>%
    pivot_longer(-crisis_lbl, names_to = "feature", values_to = "value") %>%
    filter(!is.na(value)) %>%
    mutate(feature_label = factor(feature_names[feature], levels = feature_names[feats]))
    
  sig_df <- tibble(
    feature_label = factor(feature_names[sig_feats], levels = feature_names[feats])
  )

  if (log_y) {
    d <- d %>% mutate(value = ifelse(value > 0, log10(value), NA_real_)) %>%
      filter(!is.na(value))
    ylab <- "log10(value)"
  } else {
    ylab <- "Value"
  }

  p <- ggplot(d, aes(x = crisis_lbl, y = value, fill = crisis_lbl)) +
    geom_blank()
    
  if (nrow(sig_df) > 0) {
    p <- p + geom_rect(data = sig_df, aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf), 
                       fill = "ivory", inherit.aes = FALSE)
  }
  
  p <- p +
    geom_boxplot(alpha = 0.75,
                 outlier.size = 0.4,
                 outlier.alpha = 0.3,
                 linewidth = 0.3) +
    facet_wrap(~ feature_label, scales = "free", ncol = 3) +
    scale_fill_manual(values = crisis_colors) +
    labs(title = title,
         subtitle = "Ivory background indicates |Cohen's d| >= 0.2",
         x = NULL, y = ylab) +
    eda_theme() +
    theme(legend.position = "none",
          panel.background = element_rect(fill = "transparent", colour = NA),
          axis.title = element_blank(),
          axis.text = element_text(size = 11),
          strip.text = element_text(size = 12),      # facet title
          legend.text = element_text(size = 11),
          plot.title = element_text(size = 15, face = "bold", hjust = 0.5),  # centered
          plot.subtitle = element_text(size = 12, hjust = 0.5, face = "italic"))
          
  return(p)
}
```

### 9.1 Financial ratios

```{r box-ratios, fig.height=5}
plot_box(feature_groups[["Financial ratios"]],
         "Financial ratios — boxplots by crisis status")
```

### 9.2 Growth rates

```{r box-growth, fig.height=5}
plot_box(feature_groups[["Growth rates"]],
         "Growth rates by Crisis Status")
```

### 9.3 Ratio changes

```{r box-ratio-chg, fig.height=5}
plot_box(feature_groups[["Ratio changes"]],
         "Ratio changes by Crisis Status")
```

### 9.4 Rolling volatility

```{r box-rolling, fig.height=5}
plot_box(feature_groups[["Rolling 3y std dev"]],
         "Rolling volatility by Crisis Status")
```

### 9.5 Intra-year dynamics

```{r box-intra, fig.height=7}
plot_box(feature_groups[["Intra-year dynamics"]],
         "Intra-year dynamics by Crisis Status")
```

---

## 10. Group-wise Summary Statistics (Part C)

For every feature we report the mean and median under each class, plus
their ratio (crisis / non-crisis). A ratio far from 1 signals a strong
conditional shift.

```{r summary-stats}
summary_by_crisis <- df_model %>%
  select(crisis_lbl, all_of(feat_lookup$feature)) %>%
  pivot_longer(-crisis_lbl, names_to = "feature", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(feature, crisis_lbl) %>%
  summarise(
    mean   = mean(value),
    median = median(value),
    sd     = sd(value),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = crisis_lbl,
    values_from = c(mean, median, sd),
    names_sep   = "_"
  ) %>%
  left_join(feat_lookup, by = "feature") %>%
  mutate(
    mean_diff   = `mean_Crisis`   - `mean_Non-crisis`,
    median_diff = `median_Crisis` - `median_Non-crisis`
  ) %>%
  select(group, feature,
         mean_nc   = `mean_Non-crisis`,   mean_cr   = `mean_Crisis`,
         median_nc = `median_Non-crisis`, median_cr = `median_Crisis`,
         mean_diff, median_diff) %>%
  mutate(feature = feature_names[feature]) %>%
  arrange(group, feature)
```


## 11. Stratified Subgroup Analysis (Part D)

The discrimination results in Part A are pooled across the whole panel.
A feature that only separates crisis from non-crisis for small banks, or
only for very young banks, is useful — but tells a different story than
one that discriminates everywhere. This section bins the panel by age,
employee count, and asset size, then repeats the bivariate comparison
within each bin.

## 12. Define Subgroup Bins

```{r subgroup-bins}
# Asset size quartiles (log-based labels so they stay readable)
asset_breaks <- quantile(df_model$assets,
                         probs = c(0, 0.25, 0.5, 0.75, 1),
                         na.rm = TRUE)

df_strata <- df_model %>%
  mutate(
    age_bin = cut(age,
                  breaks = c(-Inf, 10, 25, 50, Inf),
                  labels = c("3–10y", "10–25y", "25–50y", "50y+"),
                  right = FALSE),
    emp_bin = cut(num_employees,
                  breaks = c(-Inf, 20, 100, 200, Inf),
                  labels = c("<20", "20–100", "100–200", "200+"),
                  right = FALSE),
    size_bin = cut(assets,
                   breaks = unique(asset_breaks),
                   labels = c("Q1 (smallest)", "Q2", "Q3", "Q4 (largest)"),
                   include.lowest = TRUE)
  )
```

## 13. Crisis Rate by Subgroup

A structural view: which kinds of banks are at elevated risk before we
bring features into the picture.

```{r subgroup-rate-fn}
plot_crisis_rate_by_bin <- function(bin_col, title) {
  d <- df_strata %>%
    filter(!is.na(.data[[bin_col]])) %>%
    group_by(bin = .data[[bin_col]]) %>%
    summarise(
      n_total = n(),
      n_banks = n_distinct(id_rssd, na.rm = TRUE),
      n_crisis = sum(crisis_lbl == "Crisis", na.rm = TRUE),
      rate = n_crisis / n_total,
      .groups = "drop"
    )

  ggplot(d, aes(x = bin, y = rate)) +
    geom_col(fill = crisis_colors["Crisis"], width = 0.6) +
    geom_text(aes(label = paste0(percent(rate, 0.01), "\n(n = ", comma(n_banks), ")")),
              vjust = -0.2, size = 3.5, lineheight = 0.9, colour = "black") +
    scale_y_continuous(labels = percent_format(accuracy = 0.1),
                       expand = expansion(mult = c(0, 0.15))) +
    labs(title = title,
         x = NULL, y = "Crisis rate") +
    eda_theme() +
    theme(plot.title = element_text(size = 15, face = "bold"), 
          axis.title = element_text(size = 12),
          axis.text = element_text(size = 10))
}
```

### 13.1 Crisis rate by bank age

```{r rate-age, fig.height=4}
subgr1 <- plot_crisis_rate_by_bin("age_bin", "Crisis rate by bank age")
```

### 13.2 Crisis rate by employee count

```{r rate-emp, fig.height=4}
subgr2 <- plot_crisis_rate_by_bin("emp_bin", "Crisis rate by number of employees")
```


```{r subgroups}
grid.arrange(subgr1, subgr2, ncol = 1)
ggsave("../results/figures/04_crisis_rate_subgroups.png", 
       plot = grid.arrange(subgr1, subgr2, ncol = 1),
       width = 8, height = 8)
```

### 13.3 Distribution of Bank Age and Size by Crisis Status

```{r density-age-size, fig.height=7}
p_age <- ggplot(df_model, aes(x = age, fill = crisis_lbl, colour = crisis_lbl)) +
  geom_density(alpha = 0.35, linewidth = 0.4) +
  scale_fill_manual(values = crisis_colors) +
  scale_colour_manual(values = crisis_colors) +
  labs(title = "Bank Age — density by crisis status",
       x = "Bank Age (years)", y = "Density") +
  eda_theme()

p_size <- ggplot(df_model %>% filter(num_employees > 0),
                 aes(x = num_employees, fill = crisis_lbl, colour = crisis_lbl)) +
  geom_density(alpha = 0.35, linewidth = 0.4) +
  scale_fill_manual(values = crisis_colors) +
  scale_colour_manual(values = crisis_colors) +
  labs(title = "Number of Employees — density by crisis status",
       x = "Number of Employees", y = "Density") +
  eda_theme()

grid.arrange(p_age, p_size, ncol = 1)
```

## 14. Does Feature Signal Survive Inside Each Subgroup?

We pick the top features by pooled |Cohen's d| (from Part A) and
recompute the effect size *within* each subgroup bin. A heatmap shows
whether each feature discriminates uniformly or only in specific
segments.


```{r subgroup-d-helpers}
# Lightweight Cohen's d — avoids the effsize dependency inside loops.
quick_d <- function(x, y) {
  x <- x[is.finite(x)]; y <- y[is.finite(y)]
  if (length(x) < 30 || length(y) < 30) return(NA_real_)
  sdx <- sd(x); sdy <- sd(y)
  if (is.na(sdx) || is.na(sdy) || sdx == 0 || sdy == 0) return(NA_real_)
  pooled <- sqrt((sdx^2 + sdy^2) / 2)
  (mean(y) - mean(x)) / pooled
}

# Top features by absolute pooled d (from the `discrim` table built in Part A).
top_feats <- discrim %>%
  filter(!is.na(cohen_d)) %>%
  slice_max(order_by = abs_d, n = 10) %>%
  pull(feature)

compute_subgroup_d <- function(bin_col) {
  df_strata %>%
    filter(!is.na(.data[[bin_col]])) %>%
    group_by(bin = .data[[bin_col]]) %>%
    summarise(across(all_of(top_feats),
                     ~ quick_d(.x[crisis == 0], .x[crisis == 1])),
              .groups = "drop") %>%
    pivot_longer(-bin, names_to = "feature", values_to = "d") %>%
    mutate(feature_label = factor(feature_names[feature], levels = rev(feature_names[top_feats])))
}
```

```{r subgroup-d-plot-fn}
plot_subgroup_heatmap <- function(bin_col, title) {
  d <- compute_subgroup_d(bin_col)

  # Symmetric colour scale around zero so direction is readable.
  lim <- max(abs(d$d), na.rm = TRUE)

  ggplot(d, aes(x = bin, y = feature_label, fill = d)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(aes(label = ifelse(is.na(d), "—", sprintf("%.2f", d))),
              size = 3, colour = "grey15") +
    scale_fill_gradient2(low  = crisis_colors["Non-crisis"],
                         mid  = "grey95",
                         high = crisis_colors["Crisis"],
                         midpoint = 0,
                         limits = c(-lim, lim),
                         na.value = "grey90",
                         name = "Cohen's d") +
    labs(title = title,
         subtitle = "Red = higher under crisis; blue = lower under crisis",
         x = NULL, y = NULL) +
    eda_theme() +
    theme(legend.position = "right",
          panel.grid = element_blank())
}
```

### 14.1 Top features × age bin

```{r d-by-age, fig.height=6}
plot_subgroup_heatmap("age_bin",
                      "Cohen's d of top features within age bins")
```

### 14.2 Top features × employee count

```{r d-by-emp, fig.height=6}
plot_subgroup_heatmap("emp_bin",
                      "Cohen's d of top features within employee-count bins")
```

### 14.3 Top features × asset size

```{r d-by-size, fig.height=6}
plot_subgroup_heatmap("size_bin",
                      "Cohen's d of top features within asset-size quartiles")
```

---

## 15. Pre-Crisis Feature Trajectories (Part E)

The strongest case for an early-warning feature set is not that features
differ *at* a crisis event, but that they drift in a predictable
direction for several years before it. This section tracks the median
trajectory of key features in the five years preceding each crisis event
and compares it to the baseline level among non-crisis bank-years.

## 16. Build the Event-Time Panel

```{r event-time-panel}
# Features to track through time — keep the list small so each panel
# is readable. Covers capital, profitability, growth, and volatility.
traj_features <- c("asset_growth", "equity_growth", "liab_growth", "loan_growth", "deposit_growth", "roe", "roa", "ytdnetinc")

# One row per (crisis event, offset in 0..5). Offset = years *before*
# the crisis event; offset 0 is the last pre-event observation.
crisis_events <- df_model %>%
  filter(crisis == 1) %>%
  select(id_rssd, event_year = year)

trajectory <- crisis_events %>%
  crossing(offset = 0:5) %>%
  mutate(obs_year = event_year - offset) %>%
  inner_join(
    df_model %>% select(id_rssd, year, all_of(traj_features)),
    by = c("id_rssd", "obs_year" = "year")
  )

cat("Crisis events:", nrow(crisis_events), "\n")
cat("Event-time observations:", nrow(trajectory), "\n")
```

## 17. Median Trajectory with IQR Band

For each offset we summarise crisis banks' feature values; the non-crisis
baseline is the global median across all non-crisis bank-years (shown as
a dashed line).

```{r traj-summaries}
traj_long <- trajectory %>%
  pivot_longer(all_of(traj_features),
               names_to = "feature", values_to = "value") %>%
  filter(is.finite(value)) %>%
  group_by(feature) %>%
  ungroup()

traj_summary <- traj_long %>%
  group_by(feature, offset) %>%
  summarise(
    median = median(value, na.rm = TRUE),
    q25    = quantile(value, 0.25, na.rm = TRUE),
    q75    = quantile(value, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # x-axis goes from -5 (far before) to 0 (just before event)
  mutate(years_before = -offset)

# Non-crisis baseline per feature
baseline <- df_model %>%
  filter(crisis == 0) %>%
  select(all_of(traj_features)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  filter(is.finite(value)) %>%
  group_by(feature) %>%
  summarise(nc_median = median(value), .groups = "drop")
```

```{r traj-plot, fig.height=8}
traj_plot_data <- traj_summary %>%
  left_join(baseline, by = "feature") %>%
  mutate(feature_label = factor(feature_names[feature], levels = feature_names[traj_features]))

ggplot(traj_plot_data, aes(x = years_before)) +
  geom_ribbon(aes(ymin = q25, ymax = q75),
              fill = crisis_colors["Crisis"], alpha = 0.18) +
  geom_line(aes(y = median),
            colour = crisis_colors["Crisis"], linewidth = 0.7) +
  geom_point(aes(y = median),
             colour = crisis_colors["Crisis"], size = 1.6) +
  geom_hline(aes(yintercept = nc_median),
             colour = crisis_colors["Non-crisis"],
             linetype = "dashed", linewidth = 0.5) +
  facet_wrap(~ feature_label, scales = "free", ncol = 3) +
  scale_x_continuous(breaks = -5:0,
                     labels = c("t-5", "t-4", "t-3", "t-2", "t-1", "t")) +
  labs(title = "Feature trajectories in the five years before a crisis event",
       subtitle = paste0("Red line/ribbon = crisis banks (median, IQR); ",
                         "blue dashed line = non-crisis baseline median"),
       x = "Years relative to crisis event",
       y = "Feature value") +
  eda_theme()
```

## 18. Gap to Baseline at Each Offset

A single number per offset makes the drift direction explicit: for each
feature and each offset we compute (crisis median - non-crisis median),
standardised by the non-crisis IQR so values are comparable across
features.

```{r traj-gap}
nc_stats <- df_model %>%
  filter(crisis == 0) %>%
  select(all_of(traj_features)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  filter(is.finite(value)) %>%
  group_by(feature) %>%
  summarise(nc_median = median(value),
            nc_iqr    = IQR(value),
            .groups = "drop")

gap_df <- traj_summary %>%
  left_join(nc_stats, by = "feature") %>%
  mutate(std_gap = (median - nc_median) / nc_iqr) %>%
  filter(is.finite(std_gap)) %>%
  mutate(feature_label = factor(feature_names[feature], levels = feature_names[traj_features]))

ggplot(gap_df,
       aes(x = years_before, y = feature_label, fill = std_gap)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.2f", std_gap)),
            size = 3, colour = "grey15") +
  scale_fill_gradient2(low  = crisis_colors["Non-crisis"],
                       mid  = "grey95",
                       high = crisis_colors["Crisis"],
                       midpoint = 0,
                       name = "Std. gap") +
  scale_x_continuous(breaks = -5:0,
                     labels = c("t-5", "t-4", "t-3", "t-2", "t-1", "t")) +
  labs(title = "Standardised gap to non-crisis baseline by event-time",
       subtitle = "Gap = (crisis median - non-crisis median) / non-crisis IQR",
       x = "Years relative to crisis event", y = NULL) +
  eda_theme() +
  theme(panel.grid = element_blank())
```

## 19. Final Feature Selection

```{r final-selection}
train <- df_model %>% filter(year <= 2012)
feature_cols <- setdiff(names(df_model), c("id_rssd", "year", "crisis", "growth"))
numeric_features <- feature_cols[sapply(df_model[feature_cols], is.numeric)]
cat("Starting features:", length(numeric_features), "\n")
cat(paste(numeric_features, collapse = ", "), "\n")
```

```{r feature-selection-loop}
targets <- c("crisis", "growth")

for (target_var in targets) {
  cat("\n=================================================================\n")
  cat("FEATURE SELECTION FOR TARGET:", target_var, "\n")
  cat("=================================================================\n\n")
  
  d_results <- tibble(feature = character(), cohens_d = numeric(), decision = character())
  
  for (feat in numeric_features) {
    x <- train[[feat]][train[[target_var]] == 0]
    y <- train[[feat]][train[[target_var]] == 1]
    x <- x[is.finite(x)]
    y <- y[is.finite(y)]
    
    if (length(x) < 30 | length(y) < 30) {
      if (feat == "num_employees") {
        d_results <- bind_rows(d_results, tibble(feature = feat, cohens_d = NA, decision = "KEEP"))
      } else {
        d_results <- bind_rows(d_results, tibble(feature = feat, cohens_d = NA, decision = "DROPPED (insufficient data)"))
      }
      next
    }
    
    d <- tryCatch(abs(cohen.d(y, x, na.rm = TRUE)$estimate), error = function(e) NA)
    
    d_threshold <- if (target_var == "crisis") 0.2 else 0.1
    if (feat == "num_employees" || (!is.na(d) && d >= d_threshold)) {
      d_results <- bind_rows(d_results, tibble(feature = feat, cohens_d = round(d, 4), decision = "KEEP"))
    } else {
      d_results <- bind_rows(d_results, tibble(feature = feat, cohens_d = round(d, 4), decision = "DROPPED"))
    }
  }
  
  d_results <- d_results %>% arrange(desc(cohens_d))
  
  cat(paste0("=== DROPPED by Cohen's d (|d| < ", d_threshold, ") ===\n"))
  d_results %>% filter(decision != "KEEP") %>% print(n = 100)
  
  cat("\n=== KEPT after Cohen's d ===\n")
  d_results %>% filter(decision == "KEEP") %>% print(n = 100)
  
  keep_d <- d_results %>% filter(decision == "KEEP") %>% pull(feature)
  cat("\nKept:", length(keep_d), "/ Dropped:", sum(d_results$decision != "KEEP"), "\n")
  
  # VIF filter
  X_train <- train[keep_d]
  X_train <- X_train %>% mutate(across(everything(), ~replace(., !is.finite(.), NA)))
  
  # Median-impute instead of dropping rows (preserves all observations)
  X_train <- X_train %>%
    mutate(across(everything(), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))
  
  keep_vif <- keep_d
  vif_drops <- tibble(step = integer(), feature = character(), vif_value = numeric())
  step <- 0
  
  repeat {
    if (length(keep_vif) < 2) break
    
    df_vif <- X_train[keep_vif]
    df_vif$dummy_y <- rnorm(nrow(df_vif))
    
    fit <- lm(dummy_y ~ ., data = df_vif)
    vifs <- tryCatch(vif(fit), error = function(e) NULL)
    
    if (is.null(vifs)) {
      cat("VIF computation failed — likely rank deficiency. Stopping.\n")
      break
    }
    
    vifs_to_check <- vifs[names(vifs) != "num_employees"]
    if (length(vifs_to_check) == 0 || max(vifs_to_check) < 10) break
    
    step <- step + 1
    worst <- names(which.max(vifs_to_check))
    vif_drops <- bind_rows(vif_drops, tibble(
      step = step, feature = worst, vif_value = round(max(vifs_to_check), 1)
    ))
    cat("Step", step, ": drop", worst, "(VIF =", round(max(vifs_to_check), 1), ")\n")
    keep_vif <- setdiff(keep_vif, worst)
  }
  
  cat("\n=== SURVIVING features for", target_var, "===\n")
  cat(paste(keep_vif, collapse = ", "), "\n")
  cat("Final count:", length(keep_vif), "\n")
  
  # Summary table
  summary_table <- d_results %>%
    mutate(
      survived_cohens_d = decision == "KEEP",
      survived_vif = feature %in% keep_vif,
      final_status = case_when(
        !survived_cohens_d ~ "Dropped (Cohen's d)",
        !survived_vif      ~ "Dropped (VIF)",
        TRUE               ~ "Selected"
      )
    ) %>%
    select(feature, cohens_d, final_status) %>%
    arrange(final_status, desc(cohens_d))
    
  cat("\nFinal counts for", target_var, ":\n")
  summary_table %>% count(final_status) %>% print()
  
  # Save
  writeLines(keep_vif, paste0("../data/processed_data/selected_features_", target_var, ".txt"))
  
  df_model_final <- df_model %>%
    select(id_rssd, year, crisis, growth, all_of(keep_vif))
  
  write_csv(df_model_final, paste0("../data/processed_data/df_model_", target_var, ".csv"))
  cat("Saved df_model_", target_var, ".csv with", ncol(df_model_final), "columns\n")
}
```


